Recursive Solutions to First-Order Model Counting

### 2022-07-31 doc/paper/paper.tex
First draft of the complete paper

\begin{definition}[Formula]\label{def:formula}
  A \emph{formula} (called a c-theory by \citet{DBLP:conf/ijcai/BroeckTMDR11})
  is a set of clauses such that all constraints and atoms `type check' with
  respect to domains.
\end{definition}

@@

\paragraph*{Notation.}
We write $\langle\rangle$ and $\langle x \rangle$ to denote an empty list and a
list with one element $x$, respectively, and $|l|$ to denote the length of list
$l$. Let $S$ be a set of constraints or literals, $V$ a set of variables, and
$x$ either a variable or a constant. We write $S[x/V]$ to denote $S$ with all
occurrences of all variables in $V$ replaced with $x$.\footnote{Note that if
  $(X, Y)$ is a two-variable constraint, substituting a constant $c$ for $X$
  would result in $(c, Y)$, which would have to be rewritten as $(Y, c)$ to fit
  the definition of a constraint.}

@@

% Connecting FCGs with formulas
Finally, we introduce a structure that represents a solution to a (W)FOMC
problem while it is still being built. A \emph{chip} is a pair $(G, L)$, where
$G$ is an FCG, and $L$ is a list of formulas, such that $|L|$ is equal to the
number of $\star$'s in $G$. The intuition behind this definition is that $L$
contains formulas that still need to be compiled. Once a formula is compiled, it
replaced one of the $\star$'s in $G$ according to a set order. We say that an
FCG is \emph{complete} (i.e., it represents a \emph{complete solution}) if it
has no $\star$'s. Similarly, a chip is complete if its FCG is complete (or,
equivalently, the list of formulas is empty).

@@

\begin{example}
  Let $\phi = \{\, c_1, c_2, c_3 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &= (\emptyset, \{\, (Y, X) \,\}, \{\, X \mapsto \Delta^\top, Y \mapsto \Delta^\top \,\}), \\
    c_2 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
          \end{multlined}\\
    c_3 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (X, x), (Z, X), (Z, x) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Gamma \,\}).
          \end{multlined}
  \end{align*}
  Domain $\Gamma$ and with its element $x \in \Gamma$ satisfy the preconditions
  for CR\@. The rule introduces a new domain $\Gamma'$ and transforms $\phi$ to
  $\phi' = (c_1', c_2', c_3')$, where
  \begin{align*}
    c_1' &= c_1, \\
    c_2' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
           \end{multlined} \\
    c_3' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (Z, X) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Gamma' \,\}).
           \end{multlined}
  \end{align*}
\end{example}

@@

TODO: write a short informal summary. Add the full info as an appendix and
reference it.

>>>>

\begin{definition}[Formula]\label{def:formula}
  A \emph{formula} (called a c-theory by \citet{DBLP:conf/ijcai/BroeckTMDR11})
  is a set of clauses s.t.\ all constraints and atoms `type check' with respect
  to domains.
\end{definition}

@@

\paragraph*{Notation.}
We write $\langle\rangle$ and $\langle x \rangle$ to denote an empty list and a
list with one element $x$, respectively, and $|l|$ to denote the length of list
$l$. We write $\pfun$ for partial functions. Let $S$ be a set of constraints or
literals, $V$ a set of variables, and $x$ either a variable or a constant. We
write $S[x/V]$ to denote $S$ with all occurrences of all variables in $V$
replaced with $x$.\footnote{Note that if $(X, Y)$ is a two-variable constraint,
  substituting a constant $c$ for $X$ would result in $(c, Y)$, which would have
  to be rewritten as $(Y, c)$ to fit the definition of a constraint.}

@@

% Connecting FCGs with formulas
Finally, we introduce a structure that represents a solution to a (W)FOMC
problem while it is still being built. A \emph{chip} is a pair $(G, L)$, where
$G$ is an FCG, and $L$ is a list of formulas, s.t. $|L|$ is equal to the number
of $\star$'s in $G$. The intuition behind this definition is that $L$ contains
formulas that still need to be compiled. Once a formula is compiled, it replaced
one of the $\star$'s in $G$ according to a set order. We say that an FCG is
\emph{complete} (i.e., it represents a \emph{complete solution}) if it has no
$\star$'s. Similarly, a chip is complete if its FCG is complete (or,
equivalently, the list of formulas is empty).

@@

\begin{example}
  Let $\phi = \{\, c_1, c_2, c_3 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &= (\emptyset, \{\, (Y, X) \,\}, \{\, X \mapsto \Delta^\top, Y \mapsto \Delta^\top \,\}), \\
    c_2 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
          \end{multlined}\\
    c_3 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \\
            \{\, (X, x), (Z, X), (Z, x) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Gamma \,\}).
          \end{multlined}
  \end{align*}
  Domain $\Gamma$ and with its element $x \in \Gamma$ satisfy the preconditions
  for CR\@. The rule introduces a new domain $\Gamma'$ and transforms $\phi$ to
  $\phi' = (c_1', c_2', c_3')$, where
  \begin{align*}
    c_1' &= c_1, \\
    c_2' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
           \end{multlined} \\
    c_3' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (Z, X) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Gamma' \,\}).
           \end{multlined}
  \end{align*}
\end{example}

@@

Let $\Doms$ be a function that maps any clause or formula to the set of domains
used within, and let $\dom(f)$ denote the domain of function $f$.

We use (integer-valued) hash functions to efficiently discard pairs of formulas
that are too different for recursion to be established. The hash code of a
clause $c = (L, C, \delta_{c})$ (denoted by $\# c$) combines the hash codes of
the sets of constants and predicates in $c$, the numbers of positive and
negative literals, the number of inequality constraints $|C|$, and the number of
variables $|\Vars(c)|$. The hash code of a formula $\phi$ combines the hash
codes of all its clauses and is denoted by $\#\phi$.

\citet{DBLP:conf/ijcai/BroeckTMDR11} use a cache to check if a formula is
identical to one of the formulas that have already been fully compiled. In
\textsc{Crane}, to facilitate recursion, we extend the caching scheme to include
formulas that have been encountered but not fully compiled yet. Formally, we
define a \emph{cache} to be a map from integers (e.g., hash codes) to sets of
pairs of the form $(\phi, v)$, where $\phi$ is a formula, and $v$ is an FCG
node.

\begin{algorithm}[t]
  \caption{The compilation rule for $\Reff$ nodes.}\label{alg:trycache}
  \KwIn{formula $\phi$, cache $C$}
  \KwOut{a set of chips}

  \ForEach{formula and node $(\psi, v) \in C(\#\phi)$}{
    $\rho \gets \identifyRecursion{$\phi$, $\psi$}$\;
    \lIf{$\rho \ne {\normalfont \texttt{null}}$}{\Return{$\{\, (\Reff_\rho(v), \langle\rangle) \,\}$}}
  }
  \Return{$\emptyset$}\;

  \Fn{\identifyRecursion{formula $\phi$, formula $\psi$, map $\rho = \emptyset$}}{
    \lIf{$|\phi| \ne |\psi|$ {\bf or} $\#\phi \ne \#\psi$}{\Return{\normalfont\texttt{null}}}
    \lIf{$\phi = \emptyset$}{\Return{$\rho$}}
    \ForEach{clause $c \in \psi$\label{line:for1}}{
      \ForEach{clause $d \in \phi$ s.t. $\#d=\#c$\label{line:for2}}{
        \ForAll{$\gamma \in \generateMaps{$c$, $d$, $\rho$}$\label{line:generateMaps}}{
          $\rho' \gets \identifyRecursion{$\phi\setminus\{\, d \,\}$, $\psi\setminus\{\, c \,\}$, $\rho\cup\gamma$}$\;\label{line:recursion}
          \lIf{$\rho' \ne {\normalfont \texttt{null}}$}{\Return{$\rho'$}}
        }
      }
      \Return{\normalfont\texttt{null}}\;
    }
  }
\end{algorithm}

% Cache $C$ is used to partition all previously-encountered formulas based on
% their hash codes.
\Cref{alg:trycache} describes the compilation rule for creating $\Reff$ nodes.
For every formula $\psi$ in the cache s.t. $\#\psi = \#\phi$, function
\texttt{r} is called to check whether a recursive call is feasible. If it is,
\texttt{r} returns a (total) map $\rho\colon \Doms(\psi) \to \Doms(\phi)$ that
shows how $\psi$ can be transformed into $\phi$ by replacing each domain
$\Omega \in \Doms(\psi)$ with $\rho(\Omega) \in \Doms(\phi)$. Otherwise,
\texttt{r} returns \texttt{null} to signify that $\phi$ and $\psi$ are too
different for recursion to work. This happens if $\phi$ and $\psi$ (or their
subformulas explored in recursive calls) are structurally different (i.e., the
numbers of clauses or the hash codes fail to match) or if a clause of $\psi$
cannot be paired with a sufficiently similar clause of $\phi$. Function
\texttt{r} works by iterating over pairs of clauses of $\phi$ and $\psi$ that
have the same hash codes. For every pair of clauses that are sufficiently
similar, \texttt{r} calls itself on the remaining clauses until the map
$\rho\colon \Doms(\psi) \pfun \Doms(\phi)$ becomes total, and all clauses are
successfully coupled.

The compatibility of two clauses is checked by function \texttt{genMaps}. It
considers every possible bijection $\beta\colon \Vars(c) \to \Vars(d)$ and map
$\gamma\colon \Doms(c) \to \Doms(d)$ s.t.
\[
  \begin{tikzcd}
    \Vars(c) \ar[r, dashed, "\beta"] \arrow[d, swap, "\delta_c"] & \Vars(d) \ar[d, "\delta_d"] \\
    \Doms(c) \ar[r, dashed, "\gamma"] \ar[d, hookrightarrow] & \Doms(d) \ar[d, hookrightarrow] \\
    \Doms(\psi) \ar[r, swap, "\rho", "|" marking, outer sep=5pt] & \Doms(\phi).
  \end{tikzcd}
\]
commutes, and $c$ becomes equal to $d$ when its variables are replaced according
to $\beta$ and its domains replaced according to $\gamma$. The function then
returns each such $\gamma$ as soon as possible. The commutativity check also
ensures that $\rho \cup \gamma$ is possible on \cref{line:recursion} of
\cref{alg:trycache}. The resulting (partial) function $\rho \cup \gamma$ is the
unique function s.t. $\rho \cup \gamma|_{\dom(\rho)} = \rho$, and
$\rho \cup \gamma|_{\dom(\gamma)} = \gamma$. The operation is defined when
$\rho|_{\dom(\rho)\cap\dom(\gamma)} = \gamma|_{\dom(\rho)\cap\dom(\gamma)}$.
